Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-21 tex/manuscript.tex
Oppdatert svarbrev, og små justeringer i manus

It should be noted that the original analyses by \citet{Towfic2014} differed from our re-analysis in one critical manner which was clarified upon contacting the authors. Their analyses used a different preprocessing of the microarray data (GEO: GSE61901) in which two technical replicates of each biological sample was included, corresponding to the two different slots of the bead-microarray. For analyses where the two slots have been combined to provide one set of expression values per sample, they refer to \citet{Bakshi2013} in which Partek was used for batch adjustment.

@@

The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2''). The experiment consist of 30 samples in 3 batches (batch1:6/2, batch2:3/4 and batch3:9/6 treatment/control samples). ComBat was applied followed by a T-test, in order to identify differentially expressed genes. First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the T-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$). In addition, the sanity check using random numbers was performed. The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b. In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis do not indicate a huge difference. Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.

>>>>

It should be noted that the original analyses by \citet{Towfic2014} differed from our re-analysis in one critical manner which was clarified upon contacting the authors. Their analyses used a different preprocessing of the microarray data (GEO: GSE61901) which included two technical replicates per biological sample, corresponding to the two different slots of the bead-microarray. For analyses where the two slots have been combined to provide one set of expression values per sample, they refer to \citet{Bakshi2013} in which Partek was used for batch adjustment.

@@

The supporting information of the original ComBat article (\citealp{Johnson2007}) demonstrates the method on cells inhibited for the expression of the TAL1 gene compared to controls on a microarray platform (denoted ``Data set 2''). The experiment consist of 30 samples in 3 batches: the number of treatment/control samples are batch1:~6/2, batch2:~3/4, and batch3:~9/6. ComBat was applied followed by a T-test, in order to identify differentially expressed genes. First, we reproduced their analysis, including the adjustment by ComBat, but using \texttt{limma} instead of the T-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in \texttt{limma}, resulting in 377 probes ($q<0.05$). In addition, the sanity check using random numbers was performed. The distribution of P-values for different settings are shown in Figure~\ref{fig:comparison}b. In contrast to the results obtained for \citet{Towfic2014} (Figure~\ref{fig:comparison}a), the P-value distributions for the alternative analysis do not indicate a huge difference. Nevertheless, the P-values may still be somewhat deflated in the ComBat adjusted analysis.
